Predictive Modeling of Drinking Water Treatment Process Chemical Doses

### 2025-12-01 Paper Draft.Rmd
Completed draft of introduction. Started Data Overview. Created overall tentative outline.

>>>>

### Background and Motivation

[Background on Drinking Water Treatment/Process Overview]

Many different factors indicate effective chemical dose performance. Examples could include:

* chemical information about the water such as turbidity, charge, or pH
* information on the size of the particles being formed in the flocculation process (found via running tests on water samples)
* observation of filter performance at the end of the process
* total chemical byproducts produced
* output water quality

The chemical processes that determine treatment process performance are complex
and highly interconnected, therefore all of these factors and more are important
in making effective decisions about chemical dosing.

***The role of an operator in deciding water treatment plant chemical doses is to
look at a wide array of these factors and make decisions about how to adjust
chemical doses.***

Treatment plant conditions can change rapidly, so having tools
that aid in selecting chemical doses in response to changing water quality is 
crucial for operators. These chemicals are one of the primary costs of the treatment 
process, costing millions of dollars per year for a mid-size treatment plant. 
The chemical byproducts produced by excessive coagulant doses also have environmental
impacts. Therefore, this is both a financial problem impacting taxpayers, and an
environmental one.

### Project Purpose
***The goal of this project is to produce a model that serves the role of a 
treatment plant operator.*** This model would provide recommended chemical doses,
based on input water quality characteristics, and how operators have dosed 
chemicals in the past. This could prove a useful tool to operators as a "starting
point" in chemical dosing. They could see a change in water quality, retrieve
the suggested chemical dose from the model, test this dose, and adjust from there
based on other information.

## Data Overview and Cleaning

The data used for this effort is a timeseries dataset from a Colorado water treatment
plant covering XX years. This treatment plant takes its water from a reservoir, which
tends to be a more stable water source than sources such as rivers or industrial supplies.

The available data are as follows, with missing date ranges noted:

* Raw Water Data
    + pH
    + Temperature
    + Turbidity
    + Suspended Grain Size Information
    + Alkalinity
    + Hardness

* Chemical Dosing Data
    + Coagulant Dose
    + Cationic Polymer Dose

## Exploratory Analysis

## Predictor Selection

## Modeling

### Linear Regression

### [Other Models]

### Tree-Based Models
The benefit of a tree-based model in this application is that it is easily interpretable.
It could serve as a decision-making tool for a treatment plant operator:

* The operator sees a change in raw water conditions
* Obtains a recommended dose based on the decision tree
* Easily interprets the "why" of the dosing recommendation by following the logic of the decision tree
* Understands which variables were most important in deciding on a dose
